---
title: "R Notebook"
output: html_notebook
---

Source: https://melbourne.figshare.com/articles/812_simulated_expression_datasets_for_differential_co-expression_analysis/8010176

```{r}
#setwd("C:/Users/bmescher/Downloads")
simdata.FULL = readRDS("C:/Users/bmescher/Desktop/UTH/Stats Genomics/sim812.rds")
```

```{r}
length(simdata.FULL)
names(simdata.FULL[[2]]) #2nd of 812 simulations
```

```{r}
dim(simdata.FULL[[1]]$scores)
dim(simdata.FULL[[2]]$scores)
```


```{r}
names(attributes(simdata.FULL[[2]]$data))
dim(simdata.FULL[[2]]$data)
length(dimnames(simdata.FULL[[2]]$data)[[1]]) #150 named genes
length(dimnames(simdata.FULL[[2]]$data)[[2]]) #391 named samples 
#classf(simdata.FULL[[2]]$data)
head(simdata.FULL[[2]]$data[,1:5])
#simitr = i from 1 to 812
#scores = 21,609 x 15 sparse matrix (mostly 0s, some 1s)
#inputmodels = 25 length, with proportion / mean / sd for each gene
#staticnet = igraph object with the 150 node network
#data = 150 x 391 - expression for each of 150 nodes
```


https://www.bioconductor.org/packages/release/bioc/html/dcanr.html

http://127.0.0.1:28260/library/dcanr/doc/dcanr_vignette.html

```{r}
library(dcanr)
dcMethods()
```

##Load the dataset included with dcanr package
http://127.0.0.1:28260/library/dcanr/doc/dcanr_vignette.html

List 2 knockdowns using getConditionNames()
```{r}
data(sim102)
#get available conditions
getConditionNames(sim102)
```


In the bigger dataset from the actual paper, the object has 812 elements (accessed with [[i]]). simitr counts up from 1 to 1000, which identifies the run of the simulation program that was ultimately saved as one of 812 converging runs for the final dataset:
```{r}
length(simdata.FULL) #from the actual paper
names(simdata.FULL[[1]])
simdata.FULL[[1]]$simitr
simdata.FULL[[23]]$simitr
simdata.FULL[[85]]$simitr #the 85th converging data is from the 102nd iteration of the simulator
simdata.FULL[[812]]$simitr
simdata.FULL[[85]]$data[1:5,1:5]
names(simdata.FULL[[85]]$inputmodels) #list of modesls, with proportion/mean/sd for each
#simdata.FULL[[85]]$numinput #12
names(attributes(simdata.FULL[[85]]$data))
dim(simdata.FULL[[85]]$data) #150 x 406
```

```{r}
simdata.FULL[[85]]$numbimodal #2... these are the number of genes that are knocked down and have a bimodal distribution!
```


Looking at the toy dataset from the package, only a single element is in this list. It is actually the 102nd iteration from the full dataset (element 85)!
```{r}
length(sim102)
names(sim102) #in the toy dataset, you only have 1 simluation (?), inseetad of firsta accessing simdata812[[1]]
sim102$simitr
sim102$data[1:5,1:5]
names(sim102$inputmodels)
sim102$numinput #12
```

```{r}
dim(sim102$data) #150 gene rows, 406 different samples
sim102$data[1:5,400]
```

### Save data from the toy dataset
For the UME6 knock-down
```{r}
simdata <- getSimData(sim102, cond.name = 'UME6', full = FALSE)
emat <- simdata$emat
ume6_kd <- simdata$condition
print(emat[1:5, 1:5]) #149 genes and 406 samples
```

```{r}
head(ume6_kd) #whether or not UME6 was knocked down (binary 1 or 2)
```

### Compute Scores

```{r}
#apply the z-score method with Spearman correlations
z_scores <- dcScore(emat, ume6_kd, dc.method = 'zscore', cor.method = 'spearman')
print(z_scores[1:5, 1:5])
```

### Perform statistical test
Using dscore() function.
```{r}
#perform a statistical test: the z-test is selected automatically
raw_p <- dcTest(z_scores, emat, ume6_kd)
print(raw_p[1:5, 1:5])
```

### Correct for multiple hypothesis testing
Testing every pairwise combination of genes. Use FDR to adjust p values from the preiovus step.

```{r}
#adjust p-values (raw p-values from dcTest should NOT be modified)
adj_p <- dcAdjust(raw_p, f = p.adjust, method = 'fdr')
print(adj_p[1:5, 1:5])
```

### Generate differential co-expression network
Default adjusted p-value thresholds of 0.1 are applied where statistical tests are performed (to control for FDR at 0.1). Then the results that pass the threshold are graphed (edges are colored based on score: negative to pos graphed from purple to green.)
```{r}
library(igraph)

#get the differential network
dcnet <- dcNetwork(z_scores, adj_p)
plot(dcnet, vertex.label = '')
```

```{r}
#convert to an adjacency matrix
adjmat <- as_adj(dcnet, sparse = FALSE)
print(adjmat[1:5, 1:5])
```

```{r}
#convert to a data.frame
edgedf <- as_data_frame(dcnet, what = 'edges')
print(head(edgedf))
```


## Now try to repeat, but on the full dataset
```{r}
simdata <- getSimData(simdata.FULL[[85]], cond.name = 'UME6', full = FALSE)
emat <- simdata$emat
ume6_kd <- simdata$condition
z_scores <- dcScore(emat, ume6_kd, dc.method = 'zscore', cor.method = 'spearman')
raw_p <- dcTest(z_scores, emat, ume6_kd)
adj_p <- dcAdjust(raw_p, f = p.adjust, method = 'fdr')
dcnet <- dcNetwork(z_scores, adj_p)
plot(dcnet, vertex.label = '')
```


## From the paper
### Simulation setup for evaluations
Method performance was evaluated across 1000 simulations. Simulated networks with 150 nodes were sampled from the S. cerevisiae network and approximately 500 expression profiles simulated from the network, resulting in 150 × 500 expression matrix. Some simulations resulted in fewer expression profiles due to the steady-state not being solved. 

### Summary statistics of simulations





https://github.com/DavisLaboratory/dcanr/blob/master/vignettes/dcanr_evaluation_vignette.Rmd

```{r}
#run a standard pipeline with the z-score method
dcnets <- dcPipeline(sim102, dc.func = 'zscore')
#plot the source network, true differential network and inferred networks
op <- par(no.readonly = TRUE)
par(mfrow = c(2, 2))
plotSimNetwork(sim102, main = 'Regulatory network')
plotSimNetwork(sim102, what = 'association', main = 'True differential association network')
plot(dcnets$ADR1, main = 'ADR1 KD predicted network')
plot(dcnets$UME6, main = 'UME6 KD predicted network')
par(op)
```

Results from a pipeline are provided as a list of `igraph` objects representing
the inferred networks, one for each knock-down. True positive predictions are
coloured based on the colour of the knocked-down node and false positives are
coloured grey. The `plotSimNetwork` function can be used to retrieve and plot
the true differential networks and the source regulatory network.

Any additional parameters for intermediate steps in a standard pipeline can be
passed as shown in the example below. However, doing so will produce results
different from the precomputed results discussed in (Section \@ref(precomp)).



```{r}
#run a standard pipeline with the z-score method with custom params
dcnets_sp <- dcPipeline(sim102,
                        dc.func = 'zscore',
                        cor.method = 'spearman', #use Spearman's correlation
                        thresh = 0.2) #cut-off for creating the network
```

If the parameters need to be specified for each condition, the `cond.args`
argument can be used. For example, the LDGM method requires the regularisation
parameter to be specified for each condition. Since `sim102` has 2 conditions
(*ADR1*-knockdown and *UME6*-knockdown), `ldgm.lambda` needs to be specified for
each case. Alternatively, the number of edges in the DC network resulting from
each condition can be passed using `ldgm.ntarget` in `cond.args`.

```{r}
sim102_lambdas = c(0.5145742607781790267651, 0.3486682118540171959609)
dcnets_ldgm <- dcPipeline(sim102,
                          dc.func = 'ldgm',
                          cond.args = list(
                            ldgm.lambda = sim102_lambdas
                          ))
```



## Retrieving pre-computed results {#precomp}
Following evaluation of a novel method, it is generally of interest to compare
how it performs against previous methods. Re-computing predictions for the 10
methods implemented in this package can be time consuming. As such, all
simulations come packaged with results from applying a standard pipeline using
each inference method. Pre-computed results are also retrieved using the
`dcPipeline` function, with the `precomputed` argument set to `TRUE`.

```{r message=FALSE, warning=FALSE}
#retrieve results of applying all available methods
allnets <- lapply(dcMethods(), function(m) {
  dcPipeline(sim102, dc.func = m, precomputed = TRUE)
})
names(allnets) <- dcMethods() #name the results based on methods
#get the size of the UME6 KD differential network
netsizes <- lapply(allnets, function(net) {
  length(igraph::E(net$UME6))
})
print(unlist(netsizes))
```



# Evaluate a pipeline
After inferring the differential co-expression network, the final step is to
evaluate performance based on the truth. True differential networks are computed
by performing a sensitivity analysis as described in the **Bhuva et al.
(manuscript in preparation)**. Three levels of true differential networks are
generated:

1. Direct - this represents direct TF-target regulatory interactions that are
affected by the knock-down (condition)
2. Influence - this network includes upstream TF-target interactions which are
indirect but causative
3. Association - this network includes all associations in the influence network
but adds all non-causative associations that are differential when conditioned
on the knock-down

All three levels of truth are included within the simulated data. Predicted
networks can be evaluated against any of these. We recommend using the
differential association network.

Along with the true differential co-expression network, a metric is required to summarise
performance of a method. The package provides 6 different metrics which can be accessed
by calling `perfMethods()`.


```{r}
#available performance metrics
print(perfMethods())
```

Evaluation is performed using the `dcEvaluate` function. We will run this with the 
precomputed inferences as described in (Section \@ref(precomp)). The F1-measure will
be computed to assess performance.

```{r message=FALSE, warning=FALSE}
#compute the F1-measure for the prediction made by each method
f1_scores <- lapply(allnets, function (net) {
  dcEvaluate(sim102, net, truth.type = 'association', combine = TRUE)
})
print(sort(unlist(f1_scores), decreasing = TRUE))

#compute the Matthew's correlation coefficient of the z-score inference
z_mcc <- dcEvaluate(sim102, dcnets, perf.method = 'MCC')
print(z_mcc)
```



# Recreate the F1 scores...
### For the package toy data
```{r}
#retrieve results of applying all available methods
allnets <- lapply(dcMethods(), function(m) {
  dcPipeline(sim102, dc.func = m, precomputed = TRUE)
})
names(allnets) <- dcMethods() #name the results based on methods
f1_scores <- lapply(allnets, function (net) {
  dcEvaluate(sim102, net, truth.type = 'association', combine = TRUE)
})

print(unlist(f1_scores))

```

### For the same iteration in the full dataset
```{r}
allnets <- lapply(dcMethods(), function(m) {
  dcPipeline(simdata.FULL[[85]], dc.func = m, precomputed = TRUE)
})
names(allnets) <- dcMethods() #name the results based on methods
f1_scores <- lapply(allnets, function (net) {
  dcEvaluate(simdata.FULL[[85]], net, truth.type = 'association', combine = TRUE)
})
print(unlist(f1_scores))
dcMethods()
```


### Association relationships
```{r}
results <- NULL
for (i in seq(812)){
  allnets <- lapply(dcMethods(), function(m) {
    dcPipeline(simdata.FULL[[i]], dc.func = m, precomputed = TRUE)
  })
  names(allnets) <- dcMethods() #name the results based on methods
  f1_scores <- lapply(allnets, function (net) {
    dcEvaluate(simdata.FULL[[i]], net, truth.type = 'association', combine = TRUE)
  })
  results <- c(results,unlist(f1_scores))
}
length(results)
8932 / 812 
final.results.association <- matrix(results,nrow=812,byrow=TRUE)
colnames(final.results.association) <- dcMethods()
write.table(final.results.association,"C:/Users/bmescher/Desktop/UTH/Stats Genomics/finalresults_association.csv",sep=",", dec=".",row.names=FALSE)
dim(final.results.association)
head(final.results.association)
```

### Influence relationships
```{r}
results <- NULL
for (i in seq(812)){
  allnets <- lapply(dcMethods(), function(m) {
    dcPipeline(simdata.FULL[[i]], dc.func = m, precomputed = TRUE)
  })
  names(allnets) <- dcMethods() #name the results based on methods
  f1_scores <- lapply(allnets, function (net) {
    dcEvaluate(simdata.FULL[[i]], net, truth.type = 'influence', combine = TRUE)
  })
  results <- c(results,unlist(f1_scores))
}
length(results)
8932 / 812 
final.results.influence <- matrix(results,nrow=812,byrow=TRUE)
colnames(final.results.influence) <- dcMethods()
write.table(final.results.influence,"C:/Users/bmescher/Desktop/UTH/Stats Genomics/finalresults_influence.csv",sep=",", dec=".",row.names=FALSE)
dim(final.results.influence)
head(final.results.influence)
```

### Direct relationships
```{r}
results <- NULL
for (i in seq(812)){
  allnets <- lapply(dcMethods(), function(m) {
    dcPipeline(simdata.FULL[[i]], dc.func = m, precomputed = TRUE)
  })
  names(allnets) <- dcMethods() #name the results based on methods
  f1_scores <- lapply(allnets, function (net) {
    dcEvaluate(simdata.FULL[[i]], net, truth.type = 'direct', combine = TRUE)
  })
  results <- c(results,unlist(f1_scores))
}
length(results)
8932 / 812 
final.results.direct <- matrix(results,nrow=812,byrow=TRUE)
colnames(final.results.direct) <- dcMethods()
write.table(final.results.direct,"C:/Users/bmescher/Desktop/UTH/Stats Genomics/finalresults_direct.csv",sep=",", dec=".",row.names=FALSE)
dim(final.results.direct)
head(final.results.direct)
```

### Convert to long tidy data
```{r}
library(tidyr)
df <- as.data.frame(final.results.association[,t])
df['TruthType'] <- 'Association'
long.df <- df %>% pivot_longer(-TruthType,names_to='Diff. Co-Exp. Method',values_to='F1')

df <- as.data.frame(final.results.influence[,t])
df['TruthType'] <- 'Influence'
long.df <- rbind(long.df,df %>% pivot_longer(-TruthType,names_to='Diff. Co-Exp. Method',values_to='F1'))

            
df <- as.data.frame(final.results.direct[,t])
df['TruthType'] <- 'Direct'
long.df <- rbind(long.df,df %>% pivot_longer(-TruthType,names_to='Diff. Co-Exp. Method',values_to='F1'))

head(long.df)
```

### Use ggplot2 to graph
https://cmdlinetips.com/2019/02/how-to-make-grouped-boxplots-with-ggplot2/
```{r}
library(ggplot2)
long.df %>% 
  ggplot(aes(x=reorder(`Diff. Co-Exp. Method`, -1*F1, FUN = sum),y=F1, fill=TruthType)) +
  geom_boxplot() +
  theme(legend.position="bottom") +
  xlab("Diff. Co-Exp. Method") +
  theme(axis.text.x = element_text(angle = 90)) +
  ylim(c(0,1)) 
```

